---
title: "Lab 19: Pertussis and the CMI-PB project"
author: "Lauren Waters"
format: pdf
---

Pertussis is a severe lung infection also known as whooping cough. 

We will begin by investigation the number of Pertussis cases per year in the US. 

This data is available on the CDC website [here](https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html)

```{r}
#/ echo=FALSE

cdc <- data.frame(
  year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L),
         cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116)
       )
```

lets have a look at the data frame 

```{r}
head(cdc)
```
>Q1. 

I want a nice plot of the number of cases per year.

```{r}
library(ggplot2)

ggplot(cdc) + 
  aes(x=year, y=cases) + 
  geom_point() +
  geom_line()  
```

>Q2. 

```{r}
ggplot(cdc) + 
  aes(x=year, y=cases) + 
  geom_point() +
  geom_line() + 
  geom_vline(xintercept = 1946, linetype = "dashed", color = "blue") +
  geom_vline(xintercept = 1996, linetype = "dashed", color = "red") +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "pink")
```

>Q3. 

There is an increase in cases after the aP vaccine is introduced. 

We will use the **jsonlite** package to access this data. The main function in this package is `read_json()`.

```{r}
library(jsonlite)

#Suject table
subject <- read_json("http://cmi-pb.org/api/subject", simplifyVector = TRUE)
specimen <- read_json("http://cmi-pb.org/api/specimen", simplifyVector = TRUE)
titer <-  read_json("http://cmi-pb.org/api/v4/plasma_ab_titer", simplifyVector = TRUE)
```

```{r}
head(subject)
```


>Q4. 

```{r}
table(subject$infancy_vac)
```

>Q5. 

```{r}
table(subject$biological_sex)
```

>Q6. 

```{r}
table(subject$race, subject$biological_sex)
```

```{r}
library(tidyverse)
```

```{r}
today()
```

```{r}
today() - mdy("12-05-2001")
```

```{r}
time_length( today() - ymd("2001-12-05"), "years")
```
>Q8.

```{r}
subject$age <-  ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
subject$age_years <-  time_length(subject$age, "years")

head(subject)
```

```{r}
ggplot(subject) +
  aes(age_years,
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```



>Q9.

```{r}
meta <- inner_join(specimen, subject)
head(meta)
```

>Q10.

```{r}
abdata <-  inner_join(titer, meta)
head(abdata)
```

>Q. How many isotypes are we measuring for all these individuals?

```{r}
table(abdata$isotype)
```

Lets focus on one of these (IgG)

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

Boxplot of MFI_normalized vs antigen 
```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) + 
  geom_boxplot()
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col = infancy_vac) + 
  geom_boxplot()
```

Focus in on IgG to the Pertussis Toxin (PT) antigen in the 2021 dataset.

```{r}
igg.pt <- igg %>% filter(antigen == "PT", dataset=="2021_dataset")
```

```{r}
ggplot(igg.pt) +
  aes(planned_day_relative_to_boost, 
      MFI_normalised, 
      col = infancy_vac, 
      group = subject_id) +
  geom_point() + 
  geom_line() + 
  geom_vline(xintercept = 0, linetype = "dashed", col = "black") +
  geom_vline(xintercept = 14, linetype = "dashed", col = "black")
```


